Formal verification of train route planning

### 2025-06-27 paper/main.tex
coquilles

A \emph{safe step} is a global routing  coming with a network configuration $\switches$ and a direction map $d_m: \trains \to \directions$ such that:
\begin{itemize}
\item there are no critical resources between any two routes,
\item $\switches$ and $d_m(\tid{i})$ are witnesses of all transitions in $\vec{x}^\tid{i}$. 
\item for any $\tid{i} \in \trains$, any component $x_j^\tid{i}$ with $j \leq k_{\tid{i}}$ occurs only once in $\vec{x}^{\tid{i}}$,
\end{itemize}

Any global routing can be seen as the composition of safe steps and actually, that provides a scheduling algorithm. We say that a sequence of blocks indexed by trains $\vec{\bid{}}_{\tid{i} \in \trains}$ is \emph{separate} if there are no two trains $\tid{i} \neq \tid{id'}$ with $\vec{\bid{}}[\tid{i}] = \vec{\bid{}}[\tid{id'}]$. This notion induces a graph: let $G = \langle V, E\rangle$ with $V$ the set of separate positions and $E$ the set of safe steps between them.  Then, to compute global routes, we look for a path in the graph from the source to the target. The  $A^*$ algorithm we implemented is proving highly effective. This strategy has a main advantage, it is quite flexible with respect to new events. Suppose that at some point a train is blocked, we may recompute safe steps for the other trains, possibly releasing a deadlock.

@@

To each safe step $X = (\vec{x}^{\tid{i}})_{\tid{i} \in \trains}$ coming with a network configuration $\switches$ and a direction map $d_m$, just below, we define the system state $S(X) = \redTuple{\guardI}{ \emptyList}{ \emptyList}{\stateTuple{\trainSeq{}}{\regulator}{\switches}{\signals}}$, next called a \emph{safe step state}. Informally, we set turnouts to their position according to $\switches$ during the initial step  and we lock signals to red for each train's destination according to $d_m$. 
\begin{itemize}
\item for each train $\tid{i}$, $\trainSeq[\tid{i}] = \trainTuple{\tid{i}}{x_1^{\tid{i}}}{\dirStop}{\su{d(\tid{i})}{\vec{x}^{\tid{i}}}}$,
\item $R = \langle E,  \bid{i} \in \blocks \mapsto 0, W\rangle$, $E$ being defined just below,
\item $W = \{ \tuple{x_1^\tid{i}, x_1^\tid{i}, 1, \tid{i}}  \mid \tid{i} \in \trains \}$, 
\item $F = \blocks \times \directions \to \{\sigred, \siggreen \}$ defined for $\bid{i} \in \blocks, \dirFmt{d} \in \directions$ by: $F(\langle \bid{i}, \dirFmt{d}\rangle) = \sigred$ if  $\bid{i} = x_1^\tid{j}$ for some $j\in \trains$, otherwise $F(\langle \bid{i}, \dirFmt{d}\rangle)  = \siggreen$. 
\end{itemize}

Now, let us turn to the definition of $E$. For any train $\tid{i} \in \trains$, we define $A= (\emptyList)_{x \in \vec{x}^\tid{i} }$. Then, set  $E[\tid{i}] = \langle d_m(\tid{i}), A[x_1 \leftarrow \langle \incr{x_1^\tid{i}}] \rangle]$.

@@

So, we suppose we are given safe step states $S_1, \ldots, S_k$ corresponding to $X_1, \ldots, X_k$, network configuration $\switches_1, \ldots, \switches_k$ and direction maps $d_{m1}, \ldots, d_{mk}$. To compute their composition $S_1 \cdot S_2 \cdots S_k$, we proceed by induction on $k$. For $k = 1$, we return $S_1$. For $k>0$, let $S = S_1 \cdots S_{k-1}  = \langle \guardG, \bufTrain, \bufSig, \langle \Gamma, \langle E, W, T\rangle, \switches, \signals\rangle \rangle$. We set $S \cdot S_k = \langle \guardG, \bufTrain, \bufSig, \langle \Gamma', \langle E', W, T\rangle, \switches', \signals\rangle \rangle$ as follows.

First,  for any train $\tid{i} \in \trains$, we append the orders of the train in $S_k$ to the corresponding ones in $S$. %let $P_0 = S.\Gamma[\tid{i}].P$ and $P_k = S_k.\Gamma[\tid{i}].P$, then set $\Gamma'[\tid{i}] = \Gamma[\tid{i}][ P \leftarrow P_0 \cdot P_k]$. 
Second, we proceed by updates of $E$ and $\switches$. At first, we set $E' = E$ and $\switches' = \switches$. We begin to append to $E'$, for any train, all the handlers within $S_k$. To each train $\tid{i}$, we had for its first step (within $X_k$, seen within the composition): an $\auth$ order in $E'[\tid{i}]$.

Consider some turnout $\sid{i}$ such that $\sid{i} \in \throughS(x_n, d_{mk}(\tid{i}), \switches_k, x_{n+1})$ with $\vec{x}$ the route corresponding to train $\tid{i}$ in $X_k$. In other words, the turnout is crossed by train $\tid{i}$ at step $n$ of $\vec{x}$ within $X_k$. Suppose there is a step $j < k$, a train $\tid{m}$ such that $\sid{i} \in \throughS(y_m, d_{mj}(\tid{m}), \switches_j, y_{m+1})$ with $\vec{y}$ being the route of train $\tid{m}$ at step $X_j$. If $\switches_j(\sid{i}) = \switches_k(\sid{i})$, we skip this turnout. Otherwise, if $\tid{i} = \tid{m}$, call $p$ the position within $E'$ corresponding to block $x_{n+1}$, we append in $E'$ at that position the order $\auth$. Otherwise, call $p$ the position in the map $E'$ corresponding to the atoms related to the block $y_{m+1}$. We append to that position the order $\turn{\sid{i}}{d_{mk}(\sid{i})})$. In other words, we append to the last step involving the turnout the order for the new position.

Consider now a block within a route of $X_k$: $x_{n-1}^\tid{i}, x_n^\tid{i}, x_{n+1}^\tid{i}$. In case there is a route $y^\tid{m}$ within  $X_j$ such that\footnote{Again, we take the one with maximal value $j$ if there is a choice.} $y_{p-1}^\tid{m}, y_p^\tid{m}, y_{p+1}^\tid{m}$ with $x_n = y_p$ and $\tid{i} \neq \tid{m}$, we need to set a guard on the common resource $x_n = y_p$ that will be crossed by train $\tid{m}$ before train $\tid{i}$ will do it. We say that there is a conflict on $x_n$. We continue the process on $y_{p-1}^\tid{m}, y_p^\tid{m}, y_{p+1}^\tid{m}$, finding some subsequent conflicts. We do so up to the initial position. Call $n$ the number of conflicts met all along that process. Now, call $p$ the position within $E'$ corresponding to the position $x_{n-1}^\tid{i}$ and $q$ the position corresponding to $y_{p+1}^\tid{m}$. We append in $E'$ to $p$ the order $\wait{x_n}{n}$ and to $q$ the order $\incr{x_n}$.

>>>>

A \emph{safe step} is a global routing  coming with a network configuration $\switches$ and a direction map $\delta: \trains \to \directions$ such that:
\begin{itemize}
\item there are no critical resources between any two routes,
\item $\switches$ and $\delta(\tid{i})$ are witnesses of all transitions in $\vec{x}^\tid{i}$. 
\item for any $\tid{i} \in \trains$, any component $x_j^\tid{i}$ with $j \leq k_{\tid{i}}$ occurs only once in $\vec{x}^{\tid{i}}$,
\end{itemize}

Any global routing can be seen as the composition of safe steps and actually, that provides a scheduling algorithm. We say that a sequence of blocks indexed by trains $\vec{\bid{}}_{\tid{i} \in \trains}$ is \emph{separate} if there are no two trains $\tid{i} \neq \tid{j}$ with $\vec{\bid{}}[\tid{i}] = \vec{\bid{}}[\tid{j}]$. This notion induces a graph: let $G = \langle V, E\rangle$ with $V$ the set of separate positions and $E$ the set of safe steps between them.  Then, to compute global routes, we look for a path in the graph from the source to the target. The  $A^*$ algorithm we implemented is proving highly effective. This strategy has a main advantage, it is quite flexible with respect to new events. Suppose that at some point a train is blocked, we may recompute safe steps for the other trains, possibly releasing a deadlock.

@@

To each safe step $X = (\vec{x}^{\tid{i}})_{\tid{i} \in \trains}$ coming with a network configuration $\switches$ and a direction map $\delta$, just below, we define the system state $S(X) = \redTuple{\guardI}{ \emptyList}{ \emptyList}{\stateTuple{\trainSeq{}}{\regulator}{\switches}{\signals}}$, next called a \emph{safe step state}. Informally, we set turnouts to their position according to $\switches$ during the initial step  and we lock signals to red for each train's destination according to $\delta$. 
\begin{itemize}
\item for each train $\tid{i}$, $\trainSeq[\tid{i}] = \trainTuple{\tid{i}}{x_1^{\tid{i}}}{\dirStop}{\su{\delta(\tid{i})}{\vec{x}^{\tid{i}}}}$,
\item $R = \langle E,  \bid{i} \in \blocks \mapsto 0, W\rangle$, $E$ being defined just below,
\item $W = \{ \tuple{x_1^\tid{i}, x_1^\tid{i}, 1, \tid{i}}  \mid \tid{i} \in \trains \}$, 
\item $F = \blocks \times \directions \to \{\sigred, \siggreen \}$ defined for $\bid{i} \in \blocks, \dirFmt{d} \in \directions$ by: $F(\langle \bid{i}, \dirFmt{d}\rangle) = \sigred$ if  $\bid{i} = x_1^\tid{j}$ for some $j\in \trains$, otherwise $F(\langle \bid{i}, \dirFmt{d}\rangle)  = \siggreen$. 
\end{itemize}

Now, let us turn to the definition of $E$. For any train $\tid{i} \in \trains$, we define $A= (\emptyList)_{x \in \vec{x}^\tid{i} }$. Then, set  $E[\tid{i}] = \langle x_1^\tid{i}, A[x_1 \leftarrow \langle \incr{x_1^\tid{i}}] \rangle]$.

@@

So, we suppose we are given safe step states $S_1, \ldots, S_k$ corresponding to $X_1, \ldots, X_k$, network configuration $\switches_1, \ldots, \switches_k$ and direction maps $\delta_1, \ldots, \delta_k$. To compute their composition $S_1 \cdot S_2 \cdots S_k$, we proceed by induction on $k$. For $k = 1$, we return $S_1$. For $k>0$, let $S = S_1 \cdots S_{k-1}  = \langle \guardG, \bufTrain, \bufSig, \langle \Gamma, \langle E, W, T\rangle, \switches, \signals\rangle \rangle$. We set $S \cdot S_k = \langle \guardG, \bufTrain, \bufSig, \langle \Gamma', \langle E', W, T\rangle, \switches', \signals\rangle \rangle$ as follows.

First,  for any train $\tid{i} \in \trains$, we append the orders of the train in $S_k$ to the corresponding ones in $S$. %let $P_0 = S.\Gamma[\tid{i}].P$ and $P_k = S_k.\Gamma[\tid{i}].P$, then set $\Gamma'[\tid{i}] = \Gamma[\tid{i}][ P \leftarrow P_0 \cdot P_k]$. 
Second, we proceed by updates of $E$ and $\switches$. At first, we set $E' = E$ and $\switches' = \switches$. We begin to append to $E'$, for any train, all the handlers within $S_k$. To each train $\tid{i}$, we add for its first step (within $X_k$, seen within the composition): an $\auth$ order in $E'[\tid{i}]$.

Consider some turnout $\sid{i}$ such that $\sid{i} \in \throughS(x_n, \delta_k(\tid{i}), \switches_k, x_{n+1})$ with $\vec{x}$ the route corresponding to train $\tid{i}$ in $X_k$. In other words, the turnout is crossed by train $\tid{i}$ at step $n$ of $\vec{x}$ within $X_k$. Suppose there is a step $j < k$, a train $\tid{m}$ such that $\sid{i} \in \throughS(y_m, \delta_j(\tid{m}), \switches_j, y_{m+1})$ with $\vec{y}$ being the route of train $\tid{m}$ at step $X_j$. If $\switches_j(\sid{i}) = \switches_k(\sid{i})$, we skip this turnout. Otherwise, if $\tid{i} = \tid{m}$, call $p$ the position within $E'$ corresponding to block $x_{n+1}$, we append in $E'$ at that position the order $\auth$. Otherwise, call $p$ the position in the map $E'$ corresponding to the atoms related to the block $y_{m+1}$. We append to that position the order $\turn{\sid{i}}{\sigma_k(\sid{i})})$. In other words, we append to the last step involving the turnout the order for the new position.

Consider now a block within a route of $X_k$: $x_{n-1}^\tid{i}, x_n^\tid{i}, x_{n+1}^\tid{i}$. In case there is a route $y^\tid{m}$ within  $X_j$ such that\footnote{Again, we take the one with maximal value $j$ if there is a choice.} $y_{p-1}^\tid{m}, y_p^\tid{m}, y_{p+1}^\tid{m}$ with $x_n = y_p$ and $\tid{i} \neq \tid{m}$, we need to set a guard on the common resource $x_n = y_p$ that will be crossed by train $\tid{m}$ before train $\tid{i}$ will do it. We say that there is a conflict on $x_n$. We continue the process on $y_{p-1}^\tid{m}, y_p^\tid{m}, y_{p+1}^\tid{m}$, finding some subsequent conflicts. We do so up to the initial position. Call $\ell$ the number of conflicts met all along that process. Now, call $p$ the position within $E'$ corresponding to the position $x_{n-1}^\tid{i}$ and $q$ the position corresponding to $y_{p+1}^\tid{m}$. We append in $E'$ to $p$ the order $\wait{x_n}{\ell+1}$ and to $q$ the order $\incr{x_n}$.
